Financialization of Commodities: an Asset Pricing Perspective

### 2024-06-11 others/paper/custom/main.Rmd
Track 'others' folder

jel:
  - G11
  - G23
  - G13
  - Q02

keywords: 
  - Commodity futures
  - financialization
  - hedging pressure
  - financial crisis

@@

```{r, echo = FALSE, message = FALSE}
packages.v <- c("tidyr", "magrittr", "knitr", "lettercase", "tinytex", "bookdown", "kableExtra")
invisible(sapply(packages.v, require, character.only = TRUE))
options(knitr.table.format = "latex")
knitr::opts_chunk$set(echo = FALSE, results = "asis")
# tickers.tb <- readRDS(file = paste0(getwd(), "/data/tickers.rds"))
# tables.tb <- readRDS(file = paste0(getwd(), "/explore/results/tables.rds"))
tickers.tb <- readRDS(file = "../data/tickers.rds")
tables.tb <- readRDS(file = "../explore/results/tables.rds")

@@

@@

The last twenty years have seen major upheavals in the commodity markets with the onset of financialization^[Starting around 2004, with a view of commodity markets as an emerging asset class and in an effort to hedge against inflation and seek diversification benefits [@buyuksahin_speculators_2014; @singleton_investor_2013], institutional investors sent forth an unprecedented flow of capital into commodity futures markets. This phenomenon is commonly referred to as the financialization of commodities [@domanski_financial_2007].] as well as the period of the 2008 financial crisis and its aftermath. Financialization appears to have affected a number of economic sectors, including agriculture and energy, and its impact has been hotly debated in the policy, legislative, regulatory, and academic spheres. A vigorous legislative debate has led to regulatory changes accompanied by an equally intense policy and academic debate across a number of disciplines about the perceived effects on commodity prices of the unprecedented inflow of institutional funds brought into commodity futures markets by new index type investors^[See section \ref{background} for a detailled discussion.]. In response to immediate policy concerns the academic debate was initially framed rather narrowly. In this paper we take a fundamentally different empirical approach that is broader in scope. We consider the entire cross-section of actively traded commodities on futures markets and use a futures based asset pricing framework that includes factors constructed using both price and non-price attributes.

Commodity futures markets have a history of being the classic physical commodity risks shifting conduit for commodity specialists, and financialization may have altered this long standing paradigm with the entry of long only commodity index traders (CITs) with very different hedging and volume demands from traditional participants. This change could have altered the hedging ontology in these markets and our asset pricing framework is well suited to detect these more subtle effects driven by changes in the nature of traders. Factor models are well suited to isolating common driving factors and detecting co-movement and building on theory we develop a bespoke commodity-based asset pricing framework that seems well adapted to study these two issues simultaneously.

@@

We study the performance of these models on a set of twenty-four US traded commodities and six UK traded (LME exchange) base metals over the 1997/2017 period that we divide into four distinct sub-periods. The first, the pre-financialization phase, spans the 1997/2003 period; the second, the financialization phase, spans the 2004/2007 period^[Starting point based on earlier studies [@baker_financialization_2014; @christoffersen_factor_2018].]; the third, 2008/2013, encompasses both the financial crisis and the loose monetary policy regimes that followed; and the fourth spans the period starting in 2014 to present.  
Consistent with earlier studies, we observe that mean returns for the US traded commodities are sharply higher over the financialization period compared to pre-financialization and this pattern extends to UK traded metals.

The arrival of financial investors raises the question of whether they functioned as liquidity providers or liquidity demanders. Arguments have been made for both sides with @moskowitz_time_2012 arguing that financial investors provide liquidity while @Kangtaletwopremiums2017 argue that they demand liquidity with @cheng_convective_2014 showing that they in fact transferred risk to hedgers during the 2008 financial crisis. We study this issue by extending the Keynesian individual hedging pressure paradigm by incorporating aggregate CHP (\textbf{CHP}), a market wide measure of hedging pressure first considered in @hong_what_2012. Our use of this measure is motivated by the cross sectional nature of financialization [@basak_model_2016; @cheng_convective_2014; @le_futures_2017]. This paradigm implies that returns for individual commodities should be high in periods where \textbf{CHP} is low (\textbf{backwardation}) and low in periods where \textbf{CHP} is high (\textbf{contango}). Over the first period the cross sectional average returns during \textbf{backwardation} for both the US and UK commodities are positive with those for the US statistically significant, and significantly higher than during \textbf{contango}, providing broad support for an aggregate version of the Keynesian hedging pressure hypothesis. Over the second period this pattern reverses, with both the US and UK commodities having significant positive returns during \textbf{contango} with these returns being significantly higher than during \textbf{backwardation}. This change in pattern suggests that the onset of financialization altered the traditional risk shifting nature of hedger’s behaviour implied by the Keynesian hypothesis and suggests that financial investors may have been liquidity demanders during financialization leading hedgers to become liquidity providers.

@goldstein_commodity_2017 as well as @goldstein_speculation_2014 argue that the arrival of financial investors has an impact on the extent of normal backwardation in commodity futures markets. Based on their analysis we would expect our average CHP factor, which is based on backwardation measured by hedger’s positions, as well our term structure factor, which is based on backwardation measured by roll yield, to be best suited to measure the impact of financialization.  
We analyze the pricing dynamics over the different periods by first studying the time series pricing performance of the various factors on all the US traded commodities with the factors constructed using the latter set of commodities. The market based factor outperforms the other three in the first period while the CHP factor shows the greatest improvement in performance in the second period with a four-fold increase in pricing performance. The performance of the market factor is driven by agricultural commodities and metals in both the first and second period and the dramatic improvement of the CHP factor in the second period is driven by precious and precious-base metals. The best performing commodities for the market factor exhibited approximately the same level of price co-movement both before and during financialization. In contrast, the best performing commodities for the CHP factor exhibited very low pair-wise correlation in the first period, around the average for US traded commodities, while during financialization the average pairwise correlation between the top performing CHP factor commodities is higher than that for the market factor and double of the average for all the US traded commodities. This increase is due to the dramatic rise in average pairwise correlation amongst the US traded metals all of which exhibited high R^2^ with respect to the CHP factor during financialization which in turn is due to their common low level of CHP over this period. Thus the CHP factor model is able to establish the strengthening link between a non-price attribute and price dynamics in the US metal sub-sector during financialization.  
We explore this issue further by constructing the four factors with the top eight commodities for a given factor (factor picks) in each period^[See section \ref{methods} for further details.] and evaluating their performance on these selected commodities as well as on all US traded commodities. The results of this exercise strongly suggest that the hedging pressure co-movement and price co-movement link was concentrated in the metals sub-sector.

The UK traded base metals provide a useful set of assets to test how widespread were the changes in pricing dynamics brought about by the onset of financialization. To that end, we examine the performance of the four factors on the six UK metals, when constructed from each set of factor picks. The results show that there was co-movement in global metals returns in the second period and that this co-movement could be detected by our hedging pressure based model.

The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, also appear to have induced significant changes in pricing dynamics. Returns on both US and UK traded commodities fell dramatically over both the third and fourth periods with the pattern of return for the US traded commodities across \textbf{CHP} regimes reverting to the Keynesian paradigm. The pricing results indicate the emergence of a systematic factor across the entire cross-section of the US commodities as well as strong evidence of cross-market linkages.

Commodity financialization raised societal concerns, leading to legislative action intended to alter the nature and functioning of these markets. The debate was initially framed around adequacy of speculation, the burning issue of the day, and the consequent analysis focused on the more mechanical effects of financialization. With the benefit of hindsight, it appears this approach was perhaps too narrow and it now seems necessary to address financialization from a broader perspective. Commodity price dynamics appear to have altered substantially in quite different ways over the financialization and the financial crisis periods and our commodity futures based asset pricing approach seems able to provide new insights into the nature of these changes: financialization was a phenomenon endogenous to the commodity markets transmitted via the commodity futures markets and mainly concentrated in a specific sector (metals) while the crisis and its aftermath seems to have delivered an exogenous shock across the entire cross-section of global commodities. We provide an empirical complement to a new stream of theoretical studies that try to model the impact of financialization on various aspects of commodity futures markets^[@etula_broker-dealer_2013, @acharya_limits_2013, @cheng_convective_2014, @leclercq_equilibrium_2014, @sockin_informational_2015, @goldstein_speculation_2014, @EkelandSpeculationcommodityfutures2016, @goldstein_commodity_2017 as well as @le_futures_2017, one of the few empirical studies on the impact of financialization on commodity-price cross-moments.] and demonstrate that the effects of financialization extend beyond the mechanical effects induced by indexation as outlined in @basak_model_2016 as well as @tang_index_2012.

@@

The period from 1998 to 2008 saw an influx of investment into commodity index linked products most of which made its way into the commodity futures market. Exposure to commodity indexes whose goal was to track the broad movement of commodity prices could be accessed by means of exchange traded funds and exchange traded notes and at least \$100 billion of new investment flowed into these products over the 1998-2008 period with this financialization phenomenon leading to substantial legislative changes. Concerns over the consequences of financialization were behind Rule 76 FR 4752 issued by the US Commodity Futures Trading Commission (CFTC) on January 26^th^, 2011. This provision emanates from the Dodd-Frank Wall Street and Consumer Protection Act of 2010 (Title VII, Section 737) that mandates the CFTC to use position limits to restrict the flow of speculative capital into a number of commodity markets. The Rule was approved in a close 3-2 vote and the ensuing rulemaking process was extremely contentious with several commissioners (Michael Dunn and Scott D. O’Malia in particular) expressing reservations about the lack of supporting evidence and the Rule also triggering thousands of comment letters as well as a lawsuit against the CFTC spearheaded by two Wall Street trade groups, the International Swaps & Derivatives Association (ISDA) and the Securities Industries & Financial Markets Association (SIFMA)^[ISDA & SIFMA v US CFTC; Complaint, 1:11-cv-02146; December 2^nd^, 2011.].

A world-wide debate has ensued about the role of index funds in commodity markets. The first responses to the 2007/2008 crisis of escalating food and energy prices took the form of policy reports, many of which reasoned that the growth of commodity index funds came along with an influx of largely speculative capital that was responsible for driving commodity prices beyond their historic highs [@de_schutter_food_2010_1; @gilbert_speculative_2010; @herman_not_2011_1; @schumann_hunger-makers_2011; @unctad_global_2009]. The US and various european countries also became involved in this issue. Early US Senate investigations drew on pricing and trading data supplied by the CFTC as well as interviews with numerous experts including hedge fund manager Michael Masters who linked the growing presence of index investors in US wheat markets to the observed surges in both futures and cash prices [@masters_testimony_2008; @masters_accidental_2008]. This contention has since come to be known as the Masters’ hypothesis and was largely endorsed by the final US Senate report on the issue [@USSenate_Excessive_speculation_wheat_2009]. At this stage, the question thrust on the academic sphere was whether “excessive speculation” (understood as the market activity peculiar to index type investors) was linked to escalating energy and agricultural commodity prices. A number of academic studies, most of which used commodity specific causality and correlation based analysis have disproved this contention, particularly in the case of agricultural and energy commodities. @irwin_devil_2009, @sanders_adequacy_2010, @sanders_impact_2011, @sanders_new_2011, @irwin_commodity_2013, @brunetti_commodity_2014, @hamilton_effects_2015 and @bruno_financialization_2017 study the agricultural markets. @buyuksahin_speculators_2011, @tokic_speculation_2012, @fattouh_role_2013, @kilian_role_2014, @knittel_simple_2016 and @manera_modelling_2016 study the energy markets. @bohl_does_2013_2, @kim_does_2015 and @boyd_prevalence_2016 study both energy and agricultural markets while @irwin_index_2011, @irwin_financialization_2012 and @stoll_commodity_2011 study the commodity markets in general.

@@

On the one hand, futures markets have a long history of serving commodity producers to allievate their commodity price and output risks. @keynes_treatise_1930 and @hicks_value_1939 emphasise the "normal" behaviour of naturally short hedgers outsourcing their commodity risk to naturally long commodity specialist speculators. This market configuration is referred to as "backwardation" where futures is expected to trade at a discount to spot, providing speculators with incentives to take the long side. The opposite market paradigm is referred to as "contango". Working [@working_theory_1948; @working_hedging_1953] and Cootner [@cootner_returns_1960; @CootnerSpeculationhedging1967] somehow relax this "supply-of-speculative-services" [@Tilllongtermperspectivecommodity2007] approach by introducing processors and merchants, also commodity experts, on the hedgers side and relating the notions of backwardation and contango more directly to the term structure via the theory of storage where the shape of the front end of the term structure (basis) is directly related to the market price of storage.  
Financialization, on the other hand, refers to the entry of financial investors into the commodity futures markets who, for the most part, are not commodity experts. Besides, these new market participants tend to exhibit herding like behaviour, often taking massive long-only positions on the whole cross-section of indexed commodities in an attempt to hedge against inflation and/or seek further diversification benefits for portfolios with, in most cases, existing large positions across various asset classes [@brunetti_speculators_2016; @boyd_prevalence_2016; @cheng_convective_2014; @juvenal_speculation_2015; @singleton_investor_2013; @tang_index_2012]. This contrasts with the traditional approaches of Keynes and Working and in this context the issues of co-movement and aggregate market participants behaviour seem particularly relevant.  
Asset pricing techniques are well suited to study co-movement. Building on theory we develop an ad-hoc commodity based asset pricing framework that we believe is well adapted to study these two issues simultaneously and implement it over four periods of interest. The first, the pre-financialization period, starts in 1997 and is naturally followed by financialization, spanning the 2004/2007 period, and the financial crisis and its aftermath, spanning the 2008/2013 period. The last period spans the 2014/present period starting with the beginning of the tapering/end of the US Quantitative Easing program (QE).  
Using the weekly price, open interest and market position data described above we construct market portfolios of commodity nearby futures as well as mimicking portfolios for market, commercial hedging pressure, term structure and liquidity risk factors in returns. While our market portfolios are long only, equally weighted portfolios of particular commodity sets, our other mimicking portfolios for risk factors are constructed as combinations of two sub-portfolios (legs), one held long, one held short, which respective constituents are pooled on a weekly basis according to risk factor specific criterias. The weekly return for a particular mimicking portfolio is calculated as the difference between the average return for constituents of the long portfolio and that for the constituents of the short portfolio. i.e, return on the mimicking portfolio for risk factor $i$ ($f_{i}$) for week $t$:

$$r_{t}^{f_{i}}=\frac{1}{x}\sum_{j=1}^{x}r_{j,t}-\frac{1}{x}\sum_{j=n-x}^{n}r_{j,t}$$
$n\equiv$ number of commodities in the set considered for mimicking portfolio construction.  
$x = \left \lceil n \cdot s \right \rceil$  
$s\equiv$ selection threshold ($\frac{1}{3}$ here).  
$r_{j,t}=\frac{C_{j,t}}{C_{j,t-1}}-1$  
$C_{j,t}\equiv$ week $t$ close price for commodity $j$.  
$j\equiv$ commodity rank in the ordered set $Y_{t}^{f_{i}}$.  
$Y_{t}^{f_{i}}\equiv$ week $t$ ordered set of the commodities considered for mimicking portfolio construction where the ordering rule is factor specific ($f_{i}$).

@@

The same operation is repeated for the two ($p=2$) and three ($p=3$) factor model cases, where regressors include in turn all combinations of any two and three risk factors respectively, as well as for the four ($p=4$) factor model case where all the risk factor mimicking portfolios are included. Results are reported for the one-factor model case with the other results available from the authors upon request.  
For each period we also implement the analysis described above over periods of low and high aggregate CHP (\textbf{CHP}), a notion which we introduce in this study and which we believe is particularly relevant in the context of financialization where issues of co-movement and aggregate market participants' behaviour play a central role. We define \textbf{CHP} as the cross-sectional average of CHPs for a particular pool of assets. i.e, Week $t$ \textbf{CHP} for asset pool $x$:

@@

Over the second period the pattern essentially reverses. For the twenty-four US commodities over this period, eighteen have higher mean returns during phases of \textbf{backwardation} over \textbf{contango}, with the difference statistically significant for four commodities and of the remaining seven the higher mean return during \textbf{backwardation} is statistically significant for one commodity. All of the six UK metals achieved higher mean returns over \textbf{backwardation} than \textbf{contango} with the difference being statistically significant for one commodity. The equally weighted US portfolio had a mean return of 11.16% over \textbf{backwardation} while it was 26.68% during \textbf{contango}, the first statistically significant at the 5% level and the second at the 1% level, with the difference statistically significant at the 1% level. The difference was considerably more pronounced for the equally weighted UK metals portfolio with a mean return of 6.43% during \textbf{backwardation} rising to 41.22% during \textbf{contango}, with the difference also significant at the 1% level.  

>>>>

author: 
  - name: Devraj Basu
    surname: Basu
    email: devraj.basu@strath.ac.uk
    affiliation: University of Strathclyde, Glasgow, UK
  - name: Olivier Bauthéac
    surname: Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliation: University of Strathclyde, Glasgow, UK

jel: G11, G23, G13, Q02
keywords: Commodity futures, financialization, hedging pressure, financial crisis

@@

```{r, echo = FALSE, message = FALSE}
packages <- c("tidyr", "magrittr", "knitr", "lettercase", "tinytex", "bookdown", "kableExtra", "here")
invisible(sapply(packages, require, character.only = TRUE))
options(knitr.table.format = "latex")
knitr::opts_chunk$set(echo = FALSE, results = "asis")
tickers <- readRDS(file = file.path(here::here(), "data", "tickers.rds"))
tables <- readRDS(file = file.path(here::here(), "explore", "results", "tables.rds.rds"))

@@

**Disclosure**

* Devraj Basu: 
  I am a co-author of this paper and wish to state that I have no conflict of interest, financial or otherwise. I do not hold paid or unpaid positions at any organization whose goals relate to the article.  
  ![](pics/signature - Devraj.png)\

**Disclosure**

* Olivier Bauthéac:
  I am a co-author of this paper and wish to state that I have no conflict of interest, financial or otherwise. I do not hold paid or unpaid positions at any organization whose goals relate to the article.  
  ![](pics/signature - Olivier.png)\

@@

The last twenty years have seen major upheavals in the commodity markets with the onset of financialization^[Starting around 2004, with a view of commodity markets as an emerging asset class and in an effort to hedge against inflation and seek diversification benefits [@buyuksahin_speculators_2014; @singleton_investor_2013], institutional investors sent forth an unprecedented flow of capital into commodity futures markets. This phenomenon is commonly referred to as the financialization of commodities [@domanski_financial_2007].] as well as the period of the 2008 financial crisis and its aftermath. Financialization appears to have affected a number of economic sectors, including agriculture and energy, and its impact has been hotly debated in the policy, legislative, regulatory, and academic spheres. A vigorous legislative debate has led to regulatory changes accompanied by an equally intense policy and academic debate across a number of disciplines about the perceived effects on commodity prices of the unprecedented inflow of institutional funds brought into commodity futures markets by new index type investors^[See section \ref{background} for a detailled discussion.]. In response to immediate policy concerns, the academic debate was initially framed rather narrowly. In this paper we take a fundamentally different empirical approach that is broader in scope. We consider the entire cross-section of actively traded commodities on futures markets and use a futures based asset pricing framework that includes factors constructed using both price and non-price attributes.

Commodity futures markets have a history of being the classic physical commodity risks shifting conduit for commodity specialists, and financialization may have altered this long standing paradigm with the entry of long only commodity index traders (CITs) with very different hedging and volume demands from traditional participants. This change could have altered the hedging ontology in these markets and our asset pricing framework is well suited to detect these more subtle effects driven by changes in the nature of traders. Factor model techniques are well suited to isolating common driving factors and detecting co-movement and building on theory we develop a bespoke commodity-based asset pricing framework that seems well adapted to study these two issues simultaneously.

@@

We study the performance of these models on a set of twenty-four US traded commodities and six UK traded (LME exchange) base metals over the 1997/2017 period that we divide into four distinct sub-periods. The first, the pre-financialization phase, spans the 1997/2003 period; the second, the financialization phase, spans the 2004/2007 period^[Starting point based on earlier studies [@baker_financialization_2014; @christoffersen_factor_2014].]; the third, 2008/2013, encompasses both the financial crisis and the loose monetary policy regimes that followed; and the fourth spans the period starting in 2014 to present.  
Consistent with earlier studies, we observe that mean returns for the US traded commodities are sharply higher over the financialization period compared to pre-financialization and this pattern extends to UK traded metals.

The cross-sectional nature of financialization [@basak_model_2016; @cheng_financialization_2014] leads us to extend the Keynesian individual hedging pressure paradigm by incorporating aggregate CHP (\textbf{CHP}), a market wide measure of hedging pressure first considered in @hong_what_2012. This paradigm implies that returns for individual commodities should be high in periods where \textbf{CHP} is low (\textbf{backwardation}) and low in periods where \textbf{CHP} is high (\textbf{contango}). Over the first period the cross sectional average returns during \textbf{backwardation} for both the US and UK commodities are positive with those for the US statistically significant, and significantly higher than during \textbf{contango}, providing broad support for an aggregate version of the Keynesian hedging pressure hypothesis. Over the second period this pattern reverses, with both the US and UK commodities having significant positive returns during \textbf{contango} with these returns being significantly higher than during \textbf{backwardation}. This change in pattern suggests that the onset of financialization altered the traditional risk shifting nature of hedger’s behaviour, which is implied by the Keynesian hypothesis. The arrival of financial investors raises the question of whether they functioned as liquidity providers or liquidity demanders. Arguments have been made for both of these possibilities with @moskowitz_time_2012 arguing that financial investors provide liquidity while @Kangtaletwopremiums2017 argue that they demand liquidity. Our analysis of returns in periods of low and high aggregate CHP before and during financialization allows us to analyse this issue from a new perspective and suggests that financial investors may have been liquidity demanders leading hedgers to become liquidity providers.

@goldstein_commodity_2017 as well as @goldstein_speculation_2014 argue that the arrival of financial investors has an impact on the extent of normal backwardation in commodity futures markets. Based on their analysis we would expect our average CHP factor, which is based on backwardation measured by hedger’s positions, as well our term structure factor, which is based on backwardation measured by roll yield, to be best suited to measure the impact of financialization.  
We analyze the pricing dynamics over the different periods by first studying the time series pricing performance of the various factors on all the US traded commodities with the factors constructed using the latter set of commodities. The market based factor outperforms the other three in the first period and it has a similar level of performance in the second period, while the CHP factor shows the greatest improvement in performance in the second period. The performance of the market factor is driven by agricultural commodities and metals in both the first and second period and the dramatic improvement of the CHP factor in the second period is driven by precious and precious-base metals. The best performing commodities for the market factor exhibited approximately the same level of price co-movement both before and during financialization. In contrast, the best performing commodities for the CHP factor exhibited very low pair-wise correlation in the first period, around the average for US traded commodities, while during financialization the average pairwise correlation between the top performing CHP factor commodities is higher than that for the market factor and double of the average for all the US traded commodities. This increase is due to the dramatic rise in average pairwise correlation amongst the US traded metals all of which exhibited high R^2^ with respect to the CHP factor during financialization which in turn is due to their common low level of CHP over this period. Thus the CHP factor model is able to establish the strengthening link between a non-price attribute and price dynamics in the US metal sub-sector during financialization.  
We explore this issue further by constructing the four factors with the top eight commodities for a given factor (factor picks) in each period^[See section \ref{methods} for further details.] and evaluating its performance on these selected commodities as well as on all US traded commodities. The results of this exercise strongly suggest that the hedging pressure co-movement and price co-movement link was concentrated in the metals sub-sector.

The UK traded base metals provide a useful set of assets to test how widespread were the changes in pricing dynamics brought about by the onset of financialization. To that end, we examine the performance of the four factors on the six UK metals, when constructed from each set of factor picks. The resuslts show that here was co-movement in global metals returns in the second period and that this co-movement could be detected by our hedging pressure based model.

The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, also appear to have induced significant changes in pricing dynamics. Returns on both US and UK traded commodities fell dramatically over both the third and fourth periods with the pattern of return for the US traded commodities across \textbf{CHP} regimes reverting to Keynesian paradigm. The pricing results indicate the emergence of a systematic factor across the entire cross-section of the US commodities as well as strong evidence of cross-market linkages.

Financialization was an issue of such policy importance that it triggered legislative action. The debate was initially framed around adequacy of speculation, the burning issue of the day, and the consequent analysis focused on the more mechanical effects of financialization. With the benefit of hindsight, it appears this approach was perhaps too narrow and it now seems necessary to address financialization from a broader perspective. Commodity price dynamics appear to have altered substantially in quite different ways over the financialization and the financial crisis periods and our commodity futures based asset pricing approach seems able to provide new insights into the nature of these changes: financialization was a phenomenon endogenous to the commodity markets transmitted via the commodity futures markets and mainly concentrated in a specific sector (metals) while the crisis and its aftermath seems to have delivered an exogenous shock across the entire cross-section of global commodities. The initial view of financialization was that it consisted of speculative flows which had the effect of driving up and creating bubble like conditions in commodity future prices^[@de_schutter_food_2010_1; @gilbert_how_2010; @gilbert_speculative_2010; @masters_testimony_2008; @masters_accidental_2008; @herman_not_2011_1; @schumann_hunger-makers_2011; @singleton_investor_2013; @unctad_global_2009.]. However, the fundamental question of the nature of the impact of financialization across the entire cross-section of commodity futures markets has not yet been completely answered. We provide an empirical complement to a new stream of theoretical studies that try to model the impact of financialization on various aspects of commodity futures markets^[@etula_broker-dealer_2013, @acharya_limits_2013, @cheng_convective_2014, @leclercq_equilibrium_2014, @sockin_informational_2015, @goldstein_speculation_2013, @EkelandSpeculationcommodityfutures2016, @goldstein_commodity_2017.] and demonstrate that the effects of financialization extend beyond the mechanical effects induced by indexation as outlined in @basak_model_2016 as well as @tang_index_2012.

@@

The period from 1998 to 2008 saw an influx of investment into commodity index linked products most of which made its way into the commodity futures market. These commodity indexes whose goal was to track the broad movement of commodity prices could be accessed by means of exchange traded funds and exchange traded notes. At least \$100 billion of new investment flowed into these products over the 1998-2008 period with this financialization phenomenon leading to substantial legislative changes. Concerns over the consequences of financialization were behind Rule 76 FR 4752 issued by the US Commodity Futures Trading Commission (CFTC) on January 26^th^, 2011. This provision emanates from the Dodd-Frank Wall Street and Consumer Protection Act of 2010 (Title VII, Section 737) that mandates the CFTC to use position limits to restrict the flow of speculative capital into a number of commodity markets. The Rule was approved in a close 3-2 vote and the ensuing rulemaking process was extremely contentious with several commissioners (Michael Dunn and Scott D. O’Malia in particular) expressing reservations about the lack of supporting evidence and the Rule also triggering thousands of comment letters as well as a lawsuit against the CFTC spearheaded by two Wall Street trade groups, the International Swaps & Derivatives Association (ISDA) and the Securities Industries & Financial Markets Association (SIFMA)^[ISDA & SIFMA v US CFTC; Complaint, 1:11-cv-02146; December 2^nd^, 2011.].

A world-wide debate has ensued about the role of index funds in commodity markets. The first responses to the 2007/2008 crisis of escalating food and energy prices took the form of policy reports, many of which reasoned that the growth of commodity index funds came along with an influx of largely speculative capital that was responsible for driving commodity prices beyond their historic highs [@de_schutter_food_2010_1; @gilbert_speculative_2010; @herman_not_2011_1; @schumann_hunger-makers_2011; @unctad_global_2009]. The US and various european countries also became involved in this issue. Early US Senate investigations drew on pricing and trading data supplied by the CFTC as well as interviews with numerous experts including hedge fund manager Michael Masters who linked the growing presence of index investors in US wheat markets to the observed surges in both futures and cash prices [@masters_testimony_2008; @masters_accidental_2008]). This contention has since come to be known as the Masters’ hypothesis and was largely endorsed by the final US Senate report on the issue [@USSenate_Excessive_speculation_wheat_2009]. At this stage, the question thrust on the academic sphere was whether “excessive speculation” (understood as the market activity peculiar to index type investors) was linked to escalating energy and agricultural commodity prices. A number of academic studies, most of which used commodity specific causality and correlation based analysis have disproved this contention, particularly in the case of agricultural and energy commodities. @irwin_devil_2009, @sanders_adequacy_2010, @sanders_impact_2011, @sanders_new_2011, @irwin_commodity_2013, @brunetti_commodity_2014, @hamilton_effects_2015 and @bruno_financialization_2017 study the agricultural markets. @buyuksahin_speculators_2011, @tokic_speculation_2012, @fattouh_role_2013, @kilian_role_2014, @knittel_simple_2016 and @manera_modelling_2016 study the energy markets. @bohl_does_2013_2, @kim_does_2015 and @boyd_prevalence_2016 study both energy and agricultural markets while @irwin_index_2011, @irwin_financialization_2012 and @stoll_commodity_2011 study the commodity markets in general.

@@

On the one hand, futures markets have a long history of serving commodity producers to allievate their commodity price and output risks. @keynes_treatise_1930 and @hicks_value_1939 emphasise the "normal" behaviour of naturally short hedgers outsourcing their commodity risk to naturally long commodity specialist speculators. This market configuration is referred to as "backwardation" where futures is expected to trade at a discount to spot, providing speculators with incentives to take the long side. The opposite market paradigm is referred to as "contango". Working [@working_theory_1948; @working_hedging_1953] and Cootner [@cootner_returns_1960; @CootnerSpeculationhedging1967] somehow relax this "supply-of-speculative-services" [@Tilllongtermperspectivecommodity2007] approach by introducing processors and merchants, also commodity experts, on the hedgers side and relating the notions of backwardation and contango more directly to the term structure via the theory of storage where the shape of the front end of the term structure (basis) is directly related to the market price of storage.  
Financialization, on the other hand, refers to the entry of financial investors into the commodity futures markets who, for the most part, are not commodity experts. Besides, these new market participants tend to exhibit herding like behaviour, often taking massive long-only positions on the whole cross-section of indexed commodities in an attempt to hedge against inflation and/or seek further diversification benefits for portfolios with, in most cases, existing large positions across various asset classes [@brunetti_speculators_2016; @boyd_prevalence_2016; @cheng_convective_2014; @juvenal_speculation_2015; @singleton_investor_2013; @tang_index_2012]. This contrasts with the traditional approaches of Keynes and Working and in this context the issues of co-movement and aggregate market participants behaviour seem particularly relevant.  
Asset pricing techniques are well suited to study co-movement. Building on theory we develop an ad-hoc commodity based asset pricing framework that we believe is well adapted to study these two issues simultaneously and implement it over four periods of interest. The first, the pre-financialization period, starts in 1997 and is naturally followed by financialization, spanning the 2004/2007 period, and the financial crisis and its aftermath, spanning the 2008/2013 period. The last period spans the 2014/present period sarting with the beginning of the tapering/end of the US Quantitative Easing program (QE).  
Using the weekly price, open interest and market position data described above we construct market portfolios of commodity nearby futures as well as mimicking portfolios for market, commercial hedging pressure, term structure and liquidity risk factors in returns. While our market portfolios are long only, equally weighted portfolios of particular commodity sets, our other mimicking portfolios for risk factors are constructed as combinations of two sub-portfolios (legs), one held long, one held short, which respective constituents are pooled on a weekly basis according to risk factor specific criterias. The weekly return for a particular mimicking portfolio is calculated as the difference between the average return for constituents of the long portfolio and that for the constituents of the short portfolio. i.e, return on the mimicking portfolio for risk factor $i$ ($f_{i}$) for week $t$:

$$r_{t}^{f_{i}}=\frac{1}{x}\sum_{j=1}^{x}r_{j,t}-\frac{1}{x}\sum_{j=n-x}^{n}r_{j,t}$$
$n\equiv$ number of commodities in the set considered for mimicking portfolio construction.  
$x = \left \lceil n \cdot s \right \rceil$  
$s\equiv$ selection threshold ($\frac{1}{3}$ here).  
$r_{j,t}=\frac{C_{j,t}}{C_{j,t-1}}-1$  
$C_{j,t}\equiv$ week $t$ close price for commodity $j$.  
$j\equiv$ commodity rank in the ordered set $Y_{t}^{f_{i}}$.  
$Y_{t}^{f_{i}}\equiv$ week $t$ ordered set of the commodities considered for mimicking portfolio construction where the ordering rule is specific to $f_{i}$.

@@

The same operation is repeated for the two ($p=2$) and three ($p=3$) factor model cases, where regressors include in turn all combinations of any two and three risk factors respectivly, as well as for the four ($p=4$) factor model case where all the risk factor mimicking portfolios are included. Results are reported for the one-factor model case with the other results available from the authors upon request.  
For each period we also implement the analysis described above over periods of low and high aggregate CHP (\textbf{CHP}), a notion which we introduce in this study and which we believe is particularly relevant in the context of financialization where issues of co-movement and aggregate market participants' behaviour play a central role. We define \textbf{CHP} as the cross-sectional average of CHPs for a particular pool of assets. i.e, Week $t$ \textbf{CHP} for asset pool $x$:

@@

Over the second period the pattern essentially reverses. For the twenty-four US commodities over this period, eighteen have higher mean returns during phases of \textbf{backwardation} over \textbf{contango}, with the difference statistically significant for four commodities and of the remaining seven the higher mean return during \textbf{backwardation} is statistically significant for one commodity. All of the six UK metals achieved higher mean returns over \textbf{backwardation} than contango with the difference being statistically significant for one commodity. The equally weighted US portfolio had a mean return of 11.16% over \textbf{backwardation} while it was 26.68% during \textbf{contango}, the first statistically significant at the 5% level and the second at the 1% level, with the difference statistically significant at the 1% level. The difference was considerably more pronounced for the equally weighted UK metals portfolio with a mean return of 6.43% during \textbf{backwardation} rising to 41.22% during \textbf{contango}, with the difference also significant at the 1% level.  
Thus the onset of financialization seems to have engendered a change in the nature of hedger’s behaviour in the aggregate, taking it away from risk shifting, the traditional Keynesian view. This phenomenon has been noted in earlier theoretical and empirical work [@DanthineInformationfuturesprices1978, @stout_why_1998] in different context from which two competing models of behaviour have emerged. The information arbitrage model implies that hedgers may often be seeking out counterparties to trade with rather than being purely passive. This issue is also raised in both @cheng_financialization_2014 and @StulzRethinkingriskmanagement1996 who point out that hedgers may be taking a view on prices just as speculators do. In fact, by hedging away some of their risk, hedgers are able to speculate more heavily based on their disagreement against speculators regarding futures price movement [@SimsekSpeculationrisksharing2013], which fits into a heterogenous expectation theory of speculation. As @stout_why_1998 points out, this disagreement-based theory of trading is one of the main reason the public and the law disapproves the speculators as this form of trading is regarded as non-productive^[@DuffieChallengespolicytreatment2014 also discusses some of the challenges faced by a policy treatment of speculative trading motivated by differences in beliefs.]. In the second model hedgers are seen as liquidity providers as in @Kangtaletwopremiums2017. This explanation is also consistent with the nature of financialization which led to the arrival of long term, long only investors who require substantial liquidity to roll-over their positions.
